---
title: "Decision Tree Challenge"
subtitle: "Feature Importance and Categorical Variable Encoding"
format:
  html: default
  pdf: default
execute:
  echo: true
  eval: true
---

# 🌳 Decision Tree Challenge - Feature Importance and Variable Encoding

## The Decision Tree Problem 🎯

Decision trees can be helpful when searching for analysis that has interpretability and versitility. However, flaws occur when we mark categorical variables as numbers. Decision trees react as if the sequence of numbers have a meaning, which can jumble the transparency of our analysis. This is why feature importance is established. Feature importance measures the amount that each variable contributes to the improvement of prediction accuracy. This applies for all splits in the tree. This is the path to understanding which variables have the greatest significance as predictors.

::: {.panel-tabset}

### R

```{r}
#| label: load-and-model-r
#| echo: true
#| message: false
#| warning: false

# Load libraries
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(rpart))
if (!require(rpart.plot, quietly = TRUE)) {
  install.packages("rpart.plot", repos = "https://cran.rstudio.com/")
  library(rpart.plot)
}

# Load data
sales_data <- read.csv("https://raw.githubusercontent.com/flyaflya/buad442Fall2025/refs/heads/main/datasets/salesPriceData.csv")

# Prepare model data (treating zipCode as numerical)
model_data <- sales_data %>%
  select(SalePrice, LotArea, YearBuilt, GrLivArea, FullBath, HalfBath, 
         BedroomAbvGr, TotRmsAbvGrd, GarageCars, zipCode) %>%
  na.omit()

# Split data
set.seed(123)
train_indices <- sample(1:nrow(model_data), 0.8 * nrow(model_data))
train_data <- model_data[train_indices, ]
test_data <- model_data[-train_indices, ]

# Build decision tree
tree_model <- rpart(SalePrice ~ ., 
                    data = train_data,
                    method = "anova",
                    control = rpart.control(maxdepth = 3, 
                                          minsplit = 20, 
                                          minbucket = 10))

cat("Model built with", sum(tree_model$frame$var == "<leaf>"), "terminal nodes\n")
```

### Python

```{python}
#| label: load-and-model-python
#| echo: true

import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
from sklearn.tree import DecisionTreeRegressor, plot_tree
from sklearn.model_selection import train_test_split
from sklearn.metrics import mean_squared_error, r2_score
import warnings
warnings.filterwarnings('ignore')

# Load data
sales_data = pd.read_csv("https://raw.githubusercontent.com/flyaflya/buad442Fall2025/refs/heads/main/datasets/salesPriceData.csv")

# Prepare model data (treating zipCode as numerical)
model_vars = ['SalePrice', 'LotArea', 'YearBuilt', 'GrLivArea', 'FullBath', 
              'HalfBath', 'BedroomAbvGr', 'TotRmsAbvGrd', 'GarageCars', 'zipCode']
model_data = sales_data[model_vars].dropna()

# Split data
X = model_data.drop('SalePrice', axis=1)
y = model_data['SalePrice']
X_train, X_test, y_train, y_test = train_test_split(X, y, test_size=0.2, random_state=123)

# Build decision tree
tree_model = DecisionTreeRegressor(max_depth=3, 
                                  min_samples_split=20, 
                                  min_samples_leaf=10, 
                                  random_state=123)
tree_model.fit(X_train, y_train)

print(f"Model built with {tree_model.get_n_leaves()} terminal nodes")
```

:::

## Tree Visualization

::: {.panel-tabset}

### R

```{r}
#| label: visualize-tree-r
#| echo: true
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

# Visualize tree
if (require(rpart.plot, quietly = TRUE)) {
  rpart.plot(tree_model, 
             type = 2,
             extra = 101,
             fallen.leaves = TRUE,
             digits = 0,
             cex = 0.8,
             main = "Decision Tree (zipCode as Numerical)")
} else {
  plot(tree_model, uniform = TRUE, main = "Decision Tree (zipCode as Numerical)")
  text(tree_model, use.n = TRUE, all = TRUE, cex = 0.8)
}
```

### Python

```{python}
#| label: visualize-tree-python
#| echo: true
#| fig-width: 10
#| fig-height: 6

# Visualize tree
plt.figure(figsize=(10, 6))
plot_tree(tree_model, 
          feature_names=X_train.columns,
          filled=True, 
          rounded=True,
          fontsize=10,
          max_depth=3)
plt.title("Decision Tree (zipCode as Numerical)")
plt.tight_layout()
plt.show()
```

:::

## Feature Importance Analysis

::: {.panel-tabset}

### R

```{r}
#| label: feature-importance-r
#| echo: false
#| message: false
#| warning: false

# Extract and display feature importance
importance_df <- data.frame(
  Feature = names(tree_model$variable.importance),
  Importance = as.numeric(tree_model$variable.importance)
) %>%
  arrange(desc(Importance)) %>%
  mutate(Importance_Percent = round(Importance / sum(Importance) * 100, 2))

# Check zipCode ranking
zipcode_rank <- which(importance_df$Feature == "zipCode")
zipcode_importance <- importance_df$Importance_Percent[zipcode_rank]
```

```{r}
#| label: importance-plot-r
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 5

# Plot feature importance
library(ggplot2)
ggplot(importance_df, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  coord_flip() +
  labs(title = "Feature Importance (zipCode as Numerical)",
       x = "Features", y = "Importance Score") +
  theme_minimal()
```

### Python

```{python}
#| label: feature-importance-python
#| echo: false

# Extract and display feature importance
importance_df = pd.DataFrame({
    'Feature': X_train.columns,
    'Importance': tree_model.feature_importances_
}).sort_values('Importance', ascending=False)

importance_df['Importance_Percent'] = (importance_df['Importance'] * 100).round(2)

# Check zipCode ranking
zipcode_rank = importance_df[importance_df['Feature'] == 'zipCode'].index[0] + 1
zipcode_importance = importance_df[importance_df['Feature'] == 'zipCode']['Importance_Percent'].iloc[0]
```

```{python}
#| label: importance-plot-python
#| echo: false
#| fig-width: 8
#| fig-height: 5

# Plot feature importance
plt.figure(figsize=(8, 5))
plt.barh(range(len(importance_df)), importance_df['Importance'], 
         color='steelblue', alpha=0.7)
plt.yticks(range(len(importance_df)), importance_df['Feature'])
plt.xlabel('Importance Score')
plt.title('Feature Importance (zipCode as Numerical)')
plt.gca().invert_yaxis()
plt.tight_layout()
plt.show()
```



### Categorical Encoding Analysis

::: {.panel-tabset}

### R

```{r}
#| label: categorical-r
#| echo: true
#| message: false
#| warning: false


# Convert zipCode to factor (categorical)
model_data_cat <- model_data %>%
  mutate(zipCode = as.factor(zipCode))

# Split data
set.seed(123)
train_indices_cat <- sample(1:nrow(model_data_cat), 0.8 * nrow(model_data_cat))
train_data_cat <- model_data_cat[train_indices_cat, ]
test_data_cat <- model_data_cat[-train_indices_cat, ]

# Build decision tree with categorical zipCode
tree_model_cat <- rpart(SalePrice ~ ., 
                        data = train_data_cat,
                        method = "anova",
                        control = rpart.control(maxdepth = 3, 
                                              minsplit = 20, 
                                              minbucket = 10))

# Feature importance with categorical zipCode
importance_cat <- data.frame(
  Feature = names(tree_model_cat$variable.importance),
  Importance = as.numeric(tree_model_cat$variable.importance)
) %>%
  arrange(desc(Importance)) %>%
  mutate(Importance_Percent = round(Importance / sum(Importance) * 100, 2))

# Check if zipCode appears in tree
zipcode_in_tree <- "zipCode" %in% names(tree_model_cat$variable.importance)
if(zipcode_in_tree) {
  zipcode_rank_cat <- which(importance_cat$Feature == "zipCode")
}
```

### Python

```{python}
#| label: categorical-python
#| echo: true
#| include: false
# One-hot encode zipCode
import pandas as pd

# Create one-hot encoded zipCode
zipcode_encoded = pd.get_dummies(model_data['zipCode'], prefix='zipCode')
model_data_cat = pd.concat([model_data.drop('zipCode', axis=1), zipcode_encoded], axis=1)

# Split data
X_cat = model_data_cat.drop('SalePrice', axis=1)
y_cat = model_data_cat['SalePrice']
X_train_cat, X_test_cat, y_train_cat, y_test_cat = train_test_split(X_cat, y_cat, test_size=0.2, random_state=123)

# Build decision tree with one-hot encoded zipCode
tree_model_cat = DecisionTreeRegressor(max_depth=3, 
                                      min_samples_split=20, 
                                      min_samples_leaf=10, 
                                      random_state=123)
tree_model_cat.fit(X_train_cat, y_train_cat)

# Feature importance with one-hot encoded zipCode
importance_cat_df = pd.DataFrame({
    'Feature': X_train_cat.columns,
    'Importance': tree_model_cat.feature_importances_
}).sort_values('Importance', ascending=False)

importance_cat_df['Importance_Percent'] = (importance_cat_df['Importance'] * 100).round(2)

# Check zipCode features
zipcode_features = [col for col in X_train_cat.columns if col.startswith('zipCode')]
zipcode_importance = importance_cat_df[importance_cat_df['Feature'].isin(zipcode_features)]['Importance'].sum()
total_importance = importance_cat_df['Importance'].sum()
zipcode_percent = (zipcode_importance / total_importance * 100).round(2)
```

:::

### Tree Visualization: Categorical zipCode

::: {.panel-tabset}

### R

```{r}
#| label: visualize-tree-cat-r
#| echo: true
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

# Visualize tree with categorical zipCode
if (require(rpart.plot, quietly = TRUE)) {
  rpart.plot(tree_model_cat, 
             type = 2,
             extra = 101,
             fallen.leaves = TRUE,
             digits = 0,
             cex = 0.8,
             main = "Decision Tree (zipCode as Categorical)")
} else {
  plot(tree_model_cat, uniform = TRUE, main = "Decision Tree (zipCode as Categorical)")
  text(tree_model_cat, use.n = TRUE, all = TRUE, cex = 0.8)
}
```

### Python

```{python}
#| label: visualize-tree-cat-python
#| echo: true
#| fig-width: 10
#| fig-height: 6

# Visualize tree with one-hot encoded zipCode
plt.figure(figsize=(10, 6))
plot_tree(tree_model_cat, 
          feature_names=X_train_cat.columns,
          filled=True, 
          rounded=True,
          fontsize=8,
          max_depth=4)
plt.title("Decision Tree (zipCode One-Hot Encoded)")
plt.tight_layout()
plt.show()
```

:::

### Feature Importance: Categorical zipCode

::: {.panel-tabset}

### R

```{r}
#| label: importance-plot-cat-r
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 5

# Plot feature importance for categorical zipCode
library(ggplot2)
ggplot(importance_cat, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_col(fill = "darkgreen", alpha = 0.7) +
  coord_flip() +
  labs(title = "Feature Importance (zipCode as Categorical)",
       x = "Features", y = "Importance Score") +
  theme_minimal()
```

### Python

```{python}
#| label: importance-plot-cat-python
#| echo: false
#| fig-width: 8
#| fig-height: 5

# Plot feature importance for categorical zipCode
plt.figure(figsize=(8, 5))
plt.barh(range(len(importance_cat_df)), importance_cat_df['Importance'], 
         color='darkgreen', alpha=0.7)
plt.yticks(range(len(importance_cat_df)), importance_cat_df['Feature'])
plt.xlabel('Importance Score')
plt.title('Feature Importance (zipCode One-Hot Encoded)')
plt.gca().invert_yaxis()
plt.tight_layout()
plt.show()
```

## Discussion Questions for Challenge

**Your Task:** Add thoughtful narrative answers to these two questions in the Discussion Questions section of your rendered HTML site.

1. **Numerical vs Categorical Encoding:** There are four models above, two in R and two in Python. For each language, the models differ by how zip code is modelled, either as a numerical variable or as a categorical variable. Given what you know about zip codes and real estate prices, how should zip code be modelled, numerically or categorically?

2. **R vs Python Implementation Differences:** When modelling zip code as a categorical variable, the output tree and feature importance differs quite significantly between R and Python. Investigate why this is the case. Which language would you say does a better job of modelling zip code as a categorical variable? Why is this the case? Do you see any documentation suggesting the other language does a better job? If so, please provide a quote from the documentation.

**Documentation verification required:** For question 2, you must investigate the official documentation for both `rpart` (R) and `sklearn.tree.DecisionTreeRegressor` (Python) to understand why the two implementations yield vastly different results when handling categorical variables. Your analysis should be grounded in the actual documentation and technical specifications of these libraries, not AI-generated explanations. AI can be helpful in digesting the documentation, but you must verify the information is correct.

- **Clear narrative:** Tell the story of what you discovered about decision tree feature importance
- **Insightful analysis:** Focus on the most interesting differences between numerical and categorical 
- **Human insights:** Your interpretation of what the feature importance rankings actually mean (or don't mean)
- **Documentation-based analysis:** For question 2, ground your analysis in actual library documentation

### Questions to Answer for 75% Grade on Challenge

1. **Numerical vs Categorical Analysis:** Provide a clear, well-reasoned answer to question 1 about how zip codes should be modelled. Your answer should demonstrate understanding of why categorical variables need special treatment in decision trees.

### Questions to Answer for 85% Grade on Challenge

2. **R vs Python Implementation Analysis:** Provide a thorough analysis of question 2, including investigation of the official documentation for both `rpart` (R) and `sklearn.tree.DecisionTreeRegressor` (Python). Your analysis should explain the technical differences and provide a reasoned opinion about which implementation handles categorical variables better.

### Questions to Answer for 95% Grade on Challenge

3. **Professional Presentation:** Your discussion answers should be written in a professional, engaging style that would be appropriate for a business audience. Avoid technical jargon and focus on practical implications.  Use Quarto markdown linking to create a link to the discussion section from the top of the page (see https://quarto.org/docs/authoring/cross-references.html#sections).

### Questions to Answer for 100% Grade on Challenge

4. **Documentation Integration:** For question 2, include a specific quote from the official documentation of `sklearn.tree.DecisionTreeRegressor` that supports your analysis. 


**75% Grade Requirements:**

- [ ] Clear, well-reasoned answer to question 1 about numerical vs categorical encoding

**85% Grade Requirements:**

- [ ] Thorough analysis of question 2 with investigation of official documentation

**95% Grade Requirements:**

- [ ] Professional presentation style appropriate for business audience with links to the discussion section from the top of the page (see https://quarto.org/docs/authoring/cross-references.html#sections).

**100% Grade Requirements:**

- [ ] Specific quote from official documentation of `sklearn.tree.DecisionTreeRegressor` supporting your analysis

